An IFS study of Low-powered Radio Galaxies

### 2019-08-26 paper.tex
Changes made before/after sending to Martin...?
 Changes to be committed:
	modified:   paper.tex
	modified:   refs.bib

	As can be seen from the maps in Appendix \ref{sec:maps}, both NGC 612 and NGC 3100 have significant off-centre clouds of ionized gas. The latter is spatial coincident with both the radio jet and the gap in the CO ring, while former seems to no link with either the radio jet (a massive jet extending $\sim 200$\,kpc to the east and west of the galaxy centre) or the CO.

	Total ionized gas masses are derived using the prescription of \citet{Sarzi2005}. This is a very rough calculation and the values should not be used for quantitative applications. The approach follows \citet{Kim1989}, whereby
	\begin{equation}
		\left(\frac{M_\text{\ion{H}{ii}}}{M_\odot}\right) = 280 \left(\frac{D}{10\, \mathrm{Mpc}}\right)^2 \left(\frac{F(\mathrm{H\alpha})}{10^{-14} \, \mathrm{erg \, s^{-1} \, cm^{-2}}}\right) \left(\frac{10^3 \, \mathrm{cm^{-3}}}{n_\mathrm{e}}\right) \, ,
	\end{equation}
	where $M_\text{\ion{H}{ii}}$ is the total mass of \ion{H}{ii} in the galaxy, $D$ is the galaxy distance, and $F(\mathrm{H\alpha})$ is the H$\alpha$ flux measured from the spectrum created by summing spatially all spaxels within an aperture of $\frac{R_\mathrm{e}}{2}$ (chosen to exclude noisy outer regions of the observations, but still includes the bulk of the gas in system with significant spatially-extended gas). This method assumes an electron density $n_\mathrm{e} = 100 \, \mathrm{cm^{-3}}$, a temperature of $10^4$ K and case-B recombination. Like \citet{Sarzi2005}, we only claim a detection if the amplitude-to-noise ratio (A/N) is $>4$ for [\ion{O}{iii}] and $>2.5$ for H$\beta$ or H$\alpha$ lines for VIMOS and MUSE data, respectively. When these conditions are not met, upper limits are calculated using the using the 2.5 A/N level, assuming the gas has the same velocity dispersion as the stellar component.

	Our calculated gas masses are given Table \ref{tab:derivedProp} and are at the upper limit of, and possibly exceed, the typical masses observed in ETGs. Given that our sample contains particularly massive galaxies, high ionized gas masses are not unexpected, however.

	\subsection{Ionized Gas Kinematics}
		\label{subsec:GasKin}
		Map of the best-fitting kinematics (mean velocity and velocity dispersion), as found with the methods described in Section \ref{subsubsec:EmissionFit} for the 4 galaxies (IC 1459, NGC 612, NGC 1316 and NGC 3100) with detected spatially-extended emission lines are shown in Appendix \ref{sec:maps}. Due to that fact that, unlike stars, gas is dissipational, it is to be expected that given sufficient time, the ISM will always settle into a disc. These maps show that the warm component of the ISM (the ionized gas) of the Southern sample galaxies varies from largely settled discs in IC 1459 and NGC 612 to more disordered kinematics of NGC 3100 and a possible inflow in NGC 1316.

@@

		We now comment on each galaxy in turn.

		\textbf{IC 1459} has ionized gas counter-rotating with respect to the stellar KDC, and also rotating independently from the rest of the galaxy. \citet{Franx1988} claim that the rotation of the gas is aligned with that of the stars outside of the KDC, however we observe the outer stellar kinematics to be non-rotating. As the stellar disc did not survive the merger even that created the KDC, it is unlikely that the gas in IC 1459 progenitor galaxy survived. The lost gas may, however, have been re-accreted. Alternatively, the ISM may have yet another external origin.

		\textbf{NGC 612} has well aligned stellar and gaseous kinematics. The misalignment between the position angle of the stellar and gas kinematics (using the method described in Section \ref{subsubsec:KinPA}) is just $1\fdg0\pm0\fdg7$, consistent with an internal origin for the gas. This adds to the evidence that NGC 612 has not undergone a recent major merger (major mergers exhaust galaxies of their gas and destroy a stellar disc) and suggests that, the fuelling and powering of the radio jet could be a purely secular process.

%**********************************************************************
%This plot no longer exists
%**********************************************************************
		\textbf{NGC 1316} is a complex object. Since there are redshifted emission lines with respect to the rest frame of the galaxy on both sides of the galaxy, the gas is not in a settled disc and that there must be some inflow/outflows. From visual inspection of the velocity field for the gas (Fig.\,\ref{fig:MUSE_Gaskine}) it is clear that a simple exponential disc model will not be well fitted to the data. We suggest that that the large velocity peak on the south-west side of the galaxy could be an inflow towards the nucleus, almost perpendicular to the radio jet, however we acknowledge that other interpretations are possible. Better quality data (longer exposure times) and more sophisticated models are required to add certainty to this interpretation. Nevertheless, it is clear that the gas in not in a settled disc.

@@

		Firstly, we preferentially use the BPT plots for the galaxies observed with MUSE (MUSE has the required spectral range). This allows us to classify galaxies into star-forming, LINER/LIER or Seyfert 2 classes. For the emission-line fluxes derived from the VIMOS datacubes, we use the [\ion{N}{i}]/H$\beta$ versus [\ion{O}{iii}]/H$\beta$ plot of \citet[hereafter the SAURON plot]{Sarzi2010}. Unlike the BPT plots, \citet{Sarzi2010} do not define classification boundaries. Using Eq.\,\ref{eq:N_dec} we translate the [\ion{N}{ii}]/H$\alpha$ BPT plot classification boundaries of \citet{Kewley2001} and \citet{Kauffmann2003a} to the [\ion{N}{i}]/H$\beta$ SAURON plot. This allows us to classify galaxies into star-forming or LINER/Seyfert 2 classes. We also use the so-called WHaN2 plot (H$\alpha$ equivalent width versus [\ion{N}{ii}]/H$\alpha$) from \citet{CidFernandes2011}. Using Eqs.\,\ref{eq:EqW_dec} and \ref{eq:N_dec}, we transform the boundary lines into a H$\beta$ equivalent width versus [\ion{N}{i}]/H$\beta$ (WHbN1) plot for the VIMOS data. This allows us to classify galaxies into star-forming, strong AGN (Seyfert 2), weak AGN (LINER) or retired classes.

		To check if the LINER classifications are due to a central radiation source (such as an AGN) or extended sources (such as weak star formation or post-asymptotic giant branch stars; pAGB stars), we examine the H$\alpha$ and H$\beta$ flux radial profiles of our galaxies with extended detected ionized gas (see Section \ref{subsubsec:Hb}). Finally we use the mass\,--\,excitation (MEx) plot of \cite{Nyland2016}, using parameters derived from the spatially-integrated spectra within an aperture of radius 3\arcsec\ thus only probing the central region of the galaxies (see Section \ref{subsubsec:MEx}). This allows us to classify the remaining galaxies into star-forming, Seyfert 2, LINER, transition or passive classes, and subdivide LINER galaxies into those whose LINER characteristics are attributable to a central AGN (labelled as LINER-AGN in Table \ref{tab:derivedProp}) and those whose LINER characteristics are not. We expand on the final two steps in the sections below.

@@

			\begin{figure}
				\centering
				\includegraphics[width=0.73\columnwidth]{Halpha_profile.png}
				\caption{H$\alpha$ flux radial profiles derived from the MUSE datacubes. Solid red lines show $F(\mathrm{H\alpha}) \propto r^{-2}$ convolved with a Gaussian function of width equal to the mean seeing taken from the fits file headers. This accounts for the point-spread function (PSF) of the observations.\label{fig:Ha_profile_MUSE}}
				\label{fig:Ha_profile_MUSE}
			\end{figure}

@@

			As discussed above, many galaxies classified as LINERs are in fact LIERs \citep[see e.g.][]{Sarzi2005, Sarzi2010, Singh2013, Belfiore2016}, i.e.\ the source of the ionizing radiation is not concentrated in the nuclear regions of the galaxies. To check if the ionizing radiation is entirely due to an AGN throughout the host galaxy, we plot the H$\alpha$ and H$\beta$ flux radial profiles for the MUSE and VIMOS data in Figs.\,\ref{fig:Ha_profile_MUSE} and \ref{fig:Hb_profile_VIMOS}, respectively. A point source, such as an AGN, will result in a profile having a $r^{-2}$ shape, where $r$ is the distance of the bin from the galaxy centre. A shallower profile points to a spatially-extended source, presumably non-circumnuclear stellar processes such as star formation or radiation from pAGB stars, as the dominant cause of the ionization in the outer parts of the galaxy (i.e.\ LIER rather than LINER). All the galaxies with spatially-extended Balmer emission in our Southern Sample reveal a central point source as the dominant source of the ionizing photons. The H$\beta$ profile of NGC 612 has a larger scatter suggesting that stellar processes also have some impact (likely radiation from pAGB stars, as these galaxies are ETGs and therefore likely have low star-formation rates).

@@

			The mass\,--\,excitation (MEx) plot of \citet{Juneau2011} is useful to classify galaxies with intermediate excitation levels ($0.5 < \mathrm{[\ion{O}{iii}]/H\beta} < 8.0$), particularly as it does not depend on faint lines such as the [\ion{N}{I}] doublet. We use here the calibrations by \citet{Nyland2016}, including an attempt to separate LINERs powered by AGN from those powered by pAGB stars (the former have [\ion{O}{iii}] equivalent widths $>0.8$\,\AA). We use an aperture of 3\arcsec\ to plot the nuclear MEx for our Southern Sample galaxies, and the resulting diagnostic plot is shown in Fig.\,\ref{fig:MEx}.

@@

	Overall, we thus find that the warm ISM properties of our Southern Sample galaxies are consistent with those of other optically-selected ETGs. This is consistent with low-powered radio galaxies being bona-fide detections of jet-mode AGN and therefore our sample galaxies are ordinary massive ETGs experiencing a normal but short-lived active phase, presumably due to the central black holes currently accreting gas in some form.

@@

	The two transformations that we use are from [\ion{N}{ii}]/H$\alpha$ to [\ion{N}{i}]/H$\beta$ line ratios and from the H$\alpha$ to the H$\beta$ equivalent widths. In both cases we fit a linear relationship $y = Ax + B$ using the least-trimmed squares routine, \textsc{lts\_linefit}\footnote{http://www-astro.physics.ox.ac.uk/\~mxc/software/\#lts}% \label{Cappellari}}
	 of \citet{Cappellari2013} due to its robust handling of the uncertainties along both axes. These are shown in Figs.\ \ref{fig:ratio_relation} and \ref{fig:EqW_relation}, where we find
	\begin{equation}
		\frac{[\text{\ion{N}{ii}}]}{H\alpha} = \, (0.38 \pm 0.02) \frac{[\text{\ion{N}{i}}]}{H\beta} + (1.80 \pm 0.02)
	\end{equation}
	and	
	\begin{equation}
		\mathrm{EW}(H\alpha) = \, (2.92 \pm 0.03) \mathrm{EW}(H\beta) - (0.16 \pm 0.02) \, ,
			\label{eq:EqW_dec}
	\end{equation}
	respectively. Since it is sometimes observed that [\ion{N}{ii}]/H$\alpha < 1.80$, which if blindly applying the relationship found above, results in an unphysical [\ion{N}{i}]/H$\beta < 0$ (i.e.\ negative flux), we instead rearrange and use
	\begin{equation}
		\frac{[\text{\ion{N}{i}}]}{H\beta} = \, 
		\begin{cases}
			(0.38 \pm 0.02) \frac{[\text{\ion{N}{iI}}]}{H\alpha} + (1.80 \pm 0.02), & \text{if}\ \frac{[\text{\ion{N}{ii}}]}{H\alpha} < 2 \\
			(0.38 \pm 0.02) \frac{[\text{\ion{N}{ii}}]}{H\alpha}, & \text{otherwise.}
		\end{cases}
		\label{eq:N_dec}
	\end{equation}

>>>>

	As can be seen from the maps in Appendix \ref{sec:maps}, both NGC 612 and NGC 3100 have significant off-centre clouds of ionized gas. The latter is spatial coincident with both the radio jet and the gap in the CO ring, while the former seems to have no link with either the radio jet (a massive jet extending $\sim 200$\,kpc to the east and west of the galaxy centre) or the CO.

	Total ionized gas masses are derived using the prescription of \citet{Sarzi2005}. This is a very rough calculation and the values should not be used for quantitative applications. The approach follows \citet{Kim1989}, whereby
	\begin{equation}
		\left(\frac{M_\text{\ion{H}{ii}}}{M_\odot}\right) = 280 \left(\frac{D}{10\, \mathrm{Mpc}}\right)^2 \left(\frac{F(\mathrm{H\alpha})}{10^{-14} \, \mathrm{erg \, s^{-1} \, cm^{-2}}}\right) \left(\frac{10^3 \, \mathrm{cm^{-3}}}{n_\mathrm{e}}\right) \, ,
	\end{equation}
	where $M_\text{\ion{H}{ii}}$ is the total mass of \ion{H}{ii} in the galaxy, $D$ is the galaxy distance, and $F(\mathrm{H\alpha})$ is the H$\alpha$ flux measured from the spectrum created by spatially integrating across spaxels within an aperture of $\frac{R_\mathrm{e}}{2}$ (chosen to exclude noisy outer regions of the observations, but still includes the bulk of the gas in system with significant spatially-extended gas). As in \citet{Sarzi2005} we chose an electron density $n_\mathrm{e} = 100 \, \mathrm{cm^{-3}}$. This method assumes a temperature of $10^4$ K and case-B recombination \citep{Kim1989}. Like \citet{Sarzi2005}, we only claim a detection if the amplitude-to-noise ratio (A/N) is $>4$ for [\ion{O}{iii}] and $>2.5$ for H$\beta$ or H$\alpha$ lines for VIMOS and MUSE data, respectively. When these conditions are not met, upper limits are calculated using the using the 2.5 A/N level, assuming the gas has the same velocity dispersion as the stellar component.

	Our calculated gas masses are given Table \ref{tab:derivedProp} and are at the upper limit of, and possibly exceed, the typical masses observed in ETGs \citep[e.g.][]{Phillips1986, Zeilinger1996, Sarzi2005}. Given that our sample contains particularly massive galaxies, high ionized gas masses are not unexpected, however.

	\subsection{Ionized Gas Kinematics}
		\label{subsec:GasKin}
		Maps of the best-fitting kinematics (mean velocity and velocity dispersion), as found with the methods described in Section \ref{subsubsec:EmissionFit} for the 4 galaxies (IC 1459, NGC 612, NGC 1316 and NGC 3100) with detected spatially-extended emission lines are shown in Appendix \ref{sec:maps}. Due to that fact that, unlike stars, gas is dissipational, it is to be expected that given sufficient time, the ISM will always settle into a disc. These maps show that the warm component of the ISM (the ionized gas) of the Southern sample galaxies varies from largely settled discs in IC 1459 and NGC 612 to more disordered kinematics of NGC 3100 and a possible inflow in NGC 1316.

@@

		We now comment on each of these 4 galaxies in turn.

		\textbf{IC 1459} has ionized gas counter-rotating with respect to the stellar KDC, but also not aligned with the outer stars of host galaxy (i.e. beyond the KDC) since they are non-rotating: the ionized gas has a third, independent, axis of rotation. This is contrary to \citet{Franx1988} who claims that the rotation of the gas is aligned with that of the stars outside of the KDC. As the stellar disc did not survive the merger even that created the KDC, it is unlikely that the gas in IC 1459 progenitor galaxy survived. The lost gas may, however, have been re-accreted such that the gas observed today contains a fossil record of the galaxy's original disc. Alternatively, the ISM may have yet another external origin.

		\textbf{NGC 612} has well aligned stellar and gaseous kinematics. The misalignment between the position angle of the stellar and gas kinematics (using the method described in Section \ref{subsubsec:KinPA}) is just $1\fdg0\pm0\fdg7$, consistent with an internal origin for the gas. This adds to the evidence that NGC 612 has not undergone a recent major merger (major mergers exhaust galaxies of their gas and destroy a stellar disc), suggesting that the fuelling of the radio jet could be a purely secular process.

%**********************************************************************
%This plot no longer exists
%**********************************************************************
		\textbf{NGC 1316} is a complex object. The existence of redshifted emission lines with respect to the rest frame of the galaxy on both sides of the galaxy shows the gas is not in a settled disc and that there must be some inflow/outflows. From visual inspection of the velocity field for the gas (Fig.\,\ref{fig:ngc1316}) it is clear that a simple exponential disc model will not be well fitted to the data. We suggest that that the large velocity peak on the south-west side of the galaxy could be an inflow towards the nucleus, almost perpendicular to the radio jet, however we acknowledge that other interpretations are possible. Better quality data (longer exposure times and wider fields of view) coupled with more sophisticated models are required to add certainty to this interpretation. %Nevertheless, it is clear that the gas in not in a settled disc.

@@

		Firstly, we preferentially use the BPT plots for the galaxies observed with MUSE (MUSE has the required spectral range). This allows us to classify galaxies into star-forming, LINER/LIER or Seyfert 2 classes. For the emission-line fluxes derived from the VIMOS datacubes, we use the [\ion{N}{i}]/H$\beta$ versus [\ion{O}{iii}]/H$\beta$ plot of \citet[hereafter the SAURON plot]{Sarzi2010}. Unlike the BPT plots, \citet{Sarzi2010} does not define classification boundaries. Using Eq.\,\ref{eq:N_dec} we translate the [\ion{N}{ii}]/H$\alpha$ BPT plot classification boundaries of \citet{Kewley2001} and \citet{Kauffmann2003a} to the [\ion{N}{i}]/H$\beta$ SAURON plot. This allows us to classify galaxies into star-forming or LINER/Seyfert 2 classes. We also use the so-called WHaN2 plot (H$\alpha$ equivalent width versus [\ion{N}{ii}]/H$\alpha$) from \citet{CidFernandes2011}. Using Eqs.\,\ref{eq:EqW_dec} and \ref{eq:N_dec}, we transform the boundary lines into a H$\beta$ equivalent width versus [\ion{N}{i}]/H$\beta$ (WHbN1) plot for the VIMOS data. This allows us to classify galaxies into star-forming, strong AGN (Seyfert 2), weak AGN (LINER) or retired classes.

		To check if the LINER classifications are due to a central radiation source (such as an AGN) or extended sources (such as weak star formation or post-asymptotic giant branch stars; pAGB stars), we examine the H$\alpha$ and H$\beta$ flux radial profiles of our galaxies with extended detected ionized gas (see Section \ref{subsubsec:Hb}). Finally we use the mass\,--\,excitation (MEx) plot of \cite{Nyland2016}, using parameters derived from the spatially-integrated spectra within an aperture of radius 3\arcsec\ thus only probing the central region of the galaxies (see Section \ref{subsubsec:MEx}). This allows us to classify the remaining galaxies into star-forming, Seyfert 2, LINER, transition or passive classes, and subdivide LINER galaxies into those whose LINER characteristics are attributable to a central AGN (labelled as LINER-AGN) and those whose LINER characteristics are not. As the least conventional, we expand on the final two steps in the sections below with the final classifications shown in Table \ref{tab:derivedProp}.

@@

			\begin{figure}
				\centering
				\includegraphics[width=0.73\columnwidth]{Halpha_profile.png}
				\caption{H$\alpha$ flux radial profiles derived from the MUSE datacubes. Solid red lines show $F(\mathrm{H\alpha}) \propto r^{-2}$ convolved with a Gaussian function of width equal to the mean seeing taken from the fits file headers. This accounts for the PSF of the observations.\label{fig:Ha_profile_MUSE}}
				\label{fig:Ha_profile_MUSE}
			\end{figure}

@@

			As discussed above, many galaxies classified as LINERs are in fact LIERs \citep[see e.g.][]{Sarzi2005, Sarzi2010, Singh2013, Belfiore2016}, i.e.\ the source of the ionizing radiation is not concentrated in the nuclear regions of the galaxies. To check if the ionizing radiation is entirely due to an AGN throughout the host galaxy, we plot the H$\alpha$ and H$\beta$ flux radial profiles for the MUSE and VIMOS data in Figs.\,\ref{fig:Ha_profile_MUSE} and \ref{fig:Hb_profile_VIMOS}, respectively. A point source, such as an AGN, will result in a $r^{-2}$ profile (convolved with the point-spread function; PSF), where $r$ is the distance of the bin from the galaxy centre. A shallower profile points to a spatially-extended source, presumably non-circumnuclear stellar processes such as star formation or radiation from pAGB stars, as the dominant cause of the ionization in the outer parts of the galaxy (i.e.\ a LIER rather than LINER).

			The profiles are shown in Figs.\,\ref{fig:Ha_profile_MUSE} and \ref{fig:Hb_profile_VIMOS} which show that all the galaxies with spatially-extended Balmer emission in our Southern Sample reveal a central point source as the dominant source of the ionizing photons. The H$\beta$ profile of NGC 612 has a larger scatter suggesting that stellar processes also have some impact (likely radiation from pAGB stars, as these galaxies are ETGs and therefore likely have low star-formation rates).

@@

			The mass\,--\,excitation (MEx) plot of \citet{Juneau2011} is useful to classify galaxies with intermediate excitation levels ($0.5 < \mathrm{[\ion{O}{iii}]/H\beta} < 8.0$), particularly as it does not depend on faint lines such as the [\ion{N}{I}] doublet. We use here the calibrations by \citet{Nyland2016}, including an attempt to separate LINERs powered by AGN from those powered by pAGB stars (the former have [\ion{O}{iii}] equivalent widths $>0.8$\,\AA). We focus on the nuclear region by using an aperture of 3\arcsec\ placed on our Southern Sample galaxies. The resulting diagnostic plot is shown in Fig.\,\ref{fig:MEx}.

@@

	Overall, we find that the warm ISM properties of our Southern Sample galaxies are consistent with those of other optically-selected ETGs. This is consistent with low-powered radio galaxies being bona-fide detections of jet-mode AGN and therefore our sample galaxies are ordinary massive ETGs experiencing a normal but short-lived active phase, presumably due to the central black holes currently accreting gas in some form.

@@

	The two transformations that we use are from [\ion{N}{ii}]/H$\alpha$ to [\ion{N}{i}]/H$\beta$ line ratios and from the H$\alpha$ to the H$\beta$ equivalent widths. In both cases we fit a linear relationship $y = Ax + B$ using the least-trimmed squares routine, \textsc{lts\_linefit}\footnote{http://www-astro.physics.ox.ac.uk/\~mxc/software/\#lts}% \label{Cappellari}}
	 of \citet{Cappellari2013} due to its robust handling of the uncertainties along both axes. These are shown in Figs.\ \ref{fig:ratio_relation} and \ref{fig:EqW_relation}, where we find
	\begin{equation}
		\frac{[\text{\ion{N}{ii}}]}{H\alpha} = \, (0.38 \pm 0.02) \frac{[\text{\ion{N}{i}}]}{H\beta} + (1.80 \pm 0.02)
	\end{equation}
	and	
	\begin{equation}
		\mathrm{EW}(H\alpha) = \, (2.92 \pm 0.03) \mathrm{EW}(H\beta) - (0.16 \pm 0.02) \, ,
			\label{eq:EqW_dec}
	\end{equation}
	respectively. Since it is sometimes observed that [\ion{N}{ii}]/H$\alpha < 1.80$, which if blindly applying the relationship found above, results in an unphysical [\ion{N}{i}]/H$\beta < 0$ (i.e.\ negative flux), we instead use
	\begin{equation}
		\frac{[\text{\ion{N}{i}}]}{H\beta} = \, 
		\begin{cases}
			(0.38 \pm 0.02) \frac{[\text{\ion{N}{iI}}]}{H\alpha} + (1.80 \pm 0.02), & \text{if}\ \frac{[\text{\ion{N}{ii}}]}{H\alpha} < 2 \\
			(0.38 \pm 0.02) \frac{[\text{\ion{N}{ii}}]}{H\alpha}, & \text{otherwise.}
		\end{cases}
		\label{eq:N_dec}
	\end{equation}
